---
title: "Mileston-1-Luque.Rmd"
output: 
  github_document
date:  "October 09, 2021"
author: "Ana Luque"
---

Loading the packages:
```{r}
#install packages with: install.packages() if not installed already
library(datateachr)
library(tidyverse)
```
# Task 1: Choose your favorite dataset
## 1.1
Based on the description of the data sets the ones that looked more interesting and worth of data analysis were:

_1. Flow sample_
_2. Cancer sample_
_3. Vancouver trees_
_4. Steam games_

## 1.2
Exploring more information about the tibbles.
```{r}
#Exploring the Flow_sample tibble
#To know the name of the columns, number of rows and columns glimpse function can be used
glimpse(flow_sample)
#Function used to open the data in the RStudio Data Window
tibble::view(flow_sample)
# Class prints the vector of names of classes an object inherits from.
class(flow_sample)
#Returns the first 6 rows of a data frame,
head(flow_sample)
#Summary returns an overview over the different feature columns, their type and basic distribution information.
summary(flow_sample)

#Exploring the cancer_sample
glimpse(cancer_sample)
tibble::view(cancer_sample)
class(cancer_sample)
head(cancer_sample)
summary(cancer_sample)

#Exploring the Vancouver_trees
glimpse(vancouver_trees)
tibble::view(vancouver_trees)
class(vancouver_trees)
head(vancouver_trees)
summary(vancouver_trees)

#Exploring the steam_games
glimpse(steam_games)
tibble::view(steam_games)
class(steam_games)
head(steam_games)
summary(steam_games)
```

## 1.3
By analyzing the content of the tibbles I choose the following data sets
_1. Flow sample_
_2. Cancer sample_

**Flow_sample** seems a very simple data set, however it has both quantitative and qualitative variables that can be used to group the information according to the content of the rows. Also a lot of information can be obtained from it, for example knowing in which month more flow is produced.

**Cancer_sample** I liked this data set because I am pretty familiarized with it, since my research is related to bladder cancer patients. Moreover, I have more experience working with quantitative data, because exact conclusions can be obtained from it. I would like to know if there is some relationship between the characteristic of the nuclei from the breast biopsy and the diagnosis.

## 1.4
Research questions related to each data set 
**Flow sample**
_Is there a relationship with the flow and the month of the year?_

**Cancer_sample**
_Is there a relationship between the diagnosis and the area of the nuclei in the biopsy?_

**Final choice: Cancer_sample**

# Task 2: Exploring your dataset

Exploration of the data set using dplyr and ggplot to get to know more the variables in the tibble *Cancer_samples*

First of all, a brief introduction of the data set. *Cancer_samples* was obtained from the UCI Machine Learning Repository and it contains information about the cell nuclei in the biopsy of a breast cancer mass obtained with fine needle aspirate (FNA) and analysed trough machine learning using digitized imaging. 
It contains the ID of the patient and the diagnosis B: benign tumor and M= malign tumor.

It has the mean, standard error (se), and worst (max) values of each variable:

* radius (mean of distances from center to points on the perimeter)in um
* texture (standard deviation of gray-scale values)
* perimeter in um
* area in um^2
* smoothness (local variation in radius lengths)
* compactness (perimeter^2 / area - 1.0)
* concavity (severity of concave portions of the contour)
* concave points (number of concave portions of the contour)
* symmetry
* fractal dimension ("coastline approximation" - 1)

Moreover, a brief explanation of ggplot. It is a package that can be loaded in R studio used to generate different type of graphs. The first argument is always the tibble where your data set is store, then you will set the aesthetic mapping (with aes()), the first two variables are the x and the y axis, then you can add parameters like color and fill that are related to your variables. You then add on layers named geoms (like geom_point() or geom_histogram()), inside this layers you can also specify constant parameters such as width, color and fill.

Another package I used is called dplyr, is for data manipulation, allowing you to add new variables (mutate), select a set of columns by their names (select), filter some rows (filter), reduce multiple variables into a summary (summaries) etc.. 

"%in%" Is used instead of "==" 
" %>% " Is for pipping and is used to nest functions
"<-" is used to assign a function to a variable
"print" is used to print the variable in the r console

### First excercise 
To get to know better the data set I will make a bar graph that counts the amount of patients diagnose with a benign and malign tumor. 
Furthermore, I will analyze the **distribution** of the perimeter (um) of the nuclei in the different biopsies, to see how much variability the data set has and if is worth comparing it with other variables
```{r}
# A bar chart was used to count the number of patients with malign and benign tumors, this function only need the x value, since it adds bars whose height is proportional to the number of rows. 
#To add the number of patients in the graph geom_text was used. By setting vjust (the vertical justification), it is possible to move the text above or below the tops of the bars, 1.5 to put it above the bars.

Ex1.2 <- ggplot(cancer_sample,aes(diagnosis)) +
  geom_bar(fill = "orange", alpha = 0.6) +
  geom_text(aes(label = ..count..), stat = "count", vjust = 1.5)
print(Ex1.2)

## To know the distribution of the perimeter for both benign and malign tumor geom_jitter was used,that shows in what ranges the values are located and if there is any difference between the perimeter of benign tumors and malign tumors. Since one of the main factors to detect a malign tumor during pathology analysis is to look for large nuclei. The color, transparency (alpha) and width are constant, that is why it goes directly in the geom layer outside the aes
Ex1.1<- ggplot(cancer_sample, aes(diagnosis, perimeter_mean)) + 
    geom_jitter(color= "red", alpha = 0.4, width = 0.25)
print(Ex1.1)
```

### Second exercise
Explore the **relationship between 2 variables** in a plot.I wanted to know if there is a relationship between the perimeter of the nuclei and the concavity.Since larger cells (with bigger perimeter) tend to have more concave portions of the contour. I group the plot by diagnosis to see the variability within each group or if it has some distinguish characteristics. 


```{r}
#To do a scatter plot I used geom_point with the perimeter in the x axis and the concavity in the y axis. Using the shape outside aes since is a constant value 
Ex2<- ggplot(cancer_sample, aes(perimeter_mean, concavity_mean)) +
       geom_point(mapping = aes( fill = diagnosis), shape = 21)
print(Ex2)
```

### Third exercise
I will analyze the **density** of a the area (um^2) of the nuclei in the different biopsies, to see how much variability the data set has and if is worth comparing it with other variables
```{r}
# To know the density of a single variable (the area) in both benign and malign tumors I used geom_distribution, I used ..density..  to create the histogram with a density scale. Fill was used to color the inside of the histogram
Ex3.1 <- ggplot(cancer_sample, aes(area_mean)) +
   geom_histogram(fill = "green", alpha = 0.7)
print(Ex3.1)

# I wanted to know the density of the area by diagnosis, to see if the area is different for malign and benign tumors and if there is a relationship, also to see in which the is more heterogeneity. Malign tumors tend to have bigger nuclei 
#Using geom_density you don't have to specify ..density.. it does it automatically.
#Putting fill inside of the aes in the geom_density layer is used to put in a different color the distribution of the two diagnosis.
#Alpha is used for transparency of the color (when the graphs are overlapping)
Ex3.2 <- ggplot(cancer_sample, aes(area_mean)) + 
    geom_density(aes(fill = diagnosis), alpha=0.8)
print(Ex3.2)
```
### Forth excercise 
I want to know the mean values for the two diagnostic groups, to analyze which variables are clues in detecting malign and benign tumors or if there is any difference between the two groups. Moreover, the standard deviation can be used to know the variability. For that I will be making a **new tibble with a subset of my data**. 

```{r}
# First I grouped the data set by diagnosis to have a group with all the data regarding malign tumors and data regarding benign tumors.
# The summarise_at function collapses the information of the selected columns from all the rows from each group in one 
# Mean is used to obtain the mean within the values in the group
# Pivot_longer and pivot_wider are used to transpose the data to have a better and understandable table

Ex4 <- cancer_sample %>%
  group_by(diagnosis) %>%
  summarise_at(vars(texture_mean:concavity_mean), mean) %>%
               pivot_longer(cols = -diagnosis, names_to = 'Diagnosis') %>% 
               pivot_wider(names_from = diagnosis, values_from = value)
head(Ex4)
```
# Task 3: Research questions

The main objective of analyzing this data set is to determine which variables can be used to diagnose tumors by imaging analysis of nuclei from biopsies. 

1. Is there a difference within the area of the nuclei of benign and malign tumors?
2. Does malign tumors have more variability in smoothness in their nuclei than benign tumors?
3. Could it be possible to diagnose the tumor of a patience just by looking at the concavity of the nuclei?
4. Is there a relationship between the symmetry of the nuclei and the compactness in malign tumors? 

