---
title: "Milestone-2-Luque.Rmd"
output: 
  github_document
date:  "October 19, 2021"
author: "Ana Luque"
---

Loading the packages:
```{r}
#install packages with: install.packages("package") if not installed already
library(datateachr) #contains the cancer_sample data set I am working 
library(tidyverse) #packages that allows to analyze data(tibble)
library(scales) #provides the internal scaling infrastructure used by ggplot2
library(broom) #provides the infomration of a model in a tibble
library(ggpubr) #package that allows you to combine multiple plots in one graph
theme_set(theme_pubr()) #Default function theme_pubr that creates a publication ready theme
```

# Task 1: Process and summarize your data

During Milestone 2 I will be working with the data set: **Cancer_samples**. This tibble was obtained from the UCI Machine Learning Repository and it contains information about the cell nuclei in the biopsy of a breast cancer mass obtained with fine needle aspirate (FNA) and analysed trough machine learning using digitized imaging.For more information about the data set the README file can be consulted.

Data manipulation packages *(dyplr and tidyr)* will be used to explore more in depth the data set.

## 1.1

By analyzing the content of the data set four research questions where generated, deeper analyzes is needed to answer this questions. The main idea of exploring *cancer_samples* is to determine if it is possible to use Machine learning to process biopsy from cancer patience and determine if the tumor is malign or benign. This could significantly decrease the processing time it takes for the pathology department to process samples. In cancer patience time is quiet valuable, the sooner the treatment is started the better the probabilities of surviving. 

1. Is there a difference within the area of the nuclei of benign and malign tumors?
2. Does malign tumors have more variability in smoothness in their nuclei than benign tumors?
3. Could it be possible to diagnose the tumor of a patience just by looking at the concavity of the nuclei?
4. Is there a relationship between the symmetry of the nuclei and the compactness in malign tumors? 

## 1.2 

Analyzing the research questions with R studio using summarizing and graphing packages 

#### First research questions
*Is there a difference within the area of the nuclei of benign and malign tumors?*

The area of the nuclei in cancer biopsy is critical to determine is the tissue is cancerous or not, therefore it is probable that a significant difference between the area of malign and benign tumors can be seen and used to diagnose the patients in a more efficient way. 

In order to compare the areas a *summarize table was made, including the mean, standard deviation, median, minimum and maximum values of both malign and benign tumors* so it would be easier to compare them. Moreover, a box plot will be used to visually compare the mean and if there is some data that intertwine between the two groups, a scatter plot layer will be included in the graph to see the distribution of the samples. Finally the *two geom layer graphs will contain a text layer with the mean obtained from the summarize table* 

```{r}
# The data set (tibble) that I will be using is cancer_samples, before using the functions of the tidyverse the tibble that will be analyzed must be stated.
#Group_by is used to create a grouped tibble so that the rest of the functions are perform in every group, in this case in both benign and malign tumors, and the results are shown separately.
#Arrange order the rows of a data frame in ascending order.
#Summarise allows you to create new columns of summarized variables, giving one row per group. It can be used to calculate summary statistics
#n() functions will count the number of rows in a group and will give the amount of benign and malign samples
#Mutate is used to create new variables using the existing variables, in this case is used in all numeric values to round up to one decimal
Q1.1 <- cancer_sample %>%
  group_by(diagnosis) %>%
  arrange(area_mean)  %>%
  summarise(Mean_area = mean(area_mean),
            SD_area = sd(area_mean),
            Min_area = min(area_mean),
            Median_area = median(area_mean),
            Max_area = max(area_mean), n=n()) %>%
  mutate(across(where(is.numeric), ~ round(., 1)))
Q1.1

# The aesthetic specification can be in the ggplot function or in the geom layers. To create a box plot both x and y must be specified, in this case the distribution of the area of the benign and malign tumors will be plotted
#To create a box plot the function geom_plot was used, setting a pre-difined width for the box
#Geom_jitter was added in the same graph to visualize the distribution of the samples 
#Geom_text was used to include the mean obtained in the summarize table of Q1.1, hjust is used to change the alignment horizontaly
#Labs is used to include the tittle and rename the x and y axis
Q1.2 <- ggplot(cancer_sample, aes(diagnosis, area_mean)) + 
    geom_boxplot(width = 0.3) +
    geom_jitter(alpha = 0.08, width = 0.2) +
    geom_text(data = Q1.1, aes(y=Mean_area, label = round(Mean_area,2)), hjust=-1) +
    labs(title="Area of the nuclei in breast cancer tumors", y ="Area (um^2)", x = "Diagnosis")
print(Q1.2)

#Another way of seeing if the values of the ares of both groups are overlapping is using a density plot with alpha transparency
#Alpha allows you to set a transparency value to the colors of the graph, the closer to 0 is the more transparent it will be. This allows to see if the density curves are overlapping, it can be outside the aesthetic specification because it is a pre-specified, constant value.
#The filling of the graph is inside the aesthetic specification of the geom_density because it corresponds to a variable in the data set.
Q1.3 <- ggplot(cancer_sample, aes(area_mean)) + 
    geom_density(aes(fill = diagnosis), alpha=0.8)  
print(Q1.3)

```

**Analysis**

By looking at the grouped means the difference between the two groups is evident, the area of malign tumors is 211% larger than the one of benign tumors. However, analyzing the box plot it can be seen that some samples of the two groups overlap, meaning that the diagnosis of the tumor cannot be only based in the area of their nuclei. In the box plot graph it is eventide that even that benign tumors have more samples than malign tumors, the values from malign tumors are more dispersed. The minimum value of the samples diagnosed as malign is 361, which is even lower than the mean of benign tumors. Moreover there are tree out layers with high values, the maximum value is 255% higher than the mean of malign tumors. Furthermore, in the density plot is more evident that most of nuclei of benign tumors have lower values but a significant amount of malign tumors overlap with the samples diagnosed as benign. I is evident that in average the area of the nuclei between the two groups is different but it can be used as the only approach to diagnose a cancer patient.

#### Second research questions
*Does malign tumors have more variability in smoothness in their nuclei than benign tumors* 

The smoothness of the nuclei of a tumor corresponds to the local variation in radius lengths, the closer the smoothness is to one the more irregular shapes it have. According to the literature malign breast tumors tend to have more variability in the smoothness and having more variations in their radius length, since there is more diversity in their shapes, they have lost their "normal" structure. 
Looking at the *summarize variance of the two groups* it could be possible to determine if the malign tumor have more variability in their nuclei.A common method of comparing variances is F-test, however for this test the values have to be normally distributed, so Shapiro test most be perform to know the distribution.
Moreover, using *histograms with different bin widths* the variability can be seen in a graphic way. When the sample has higher amount of taller bars across its range, far away from the mean.

```{r}
# Grouping by diagnosis and using the summarise function it is possible to determine the mean, standard deviation and variance of all the samples visualizing the general results for the malign and benign tumors. The var R function computes the sample variance of a numeric input, in this case the smoothness.The variance is a measure of dispersion that represents the variability of a series of data with respect to its medium.The standard deviation is the square root of the variance, so it also give us an idea of the variability
Q2.1 <- cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(Mean_smoothness = mean(smoothness_mean),
            Mean_sesmoothness = mean(smoothness_se),
            SD_smoothness = sd(smoothness_mean), 
            Var_smoothness= var(smoothness_mean))
head(Q2.1)

#First I separated the two diagnose groups in two different tibbles using the filter function, so every row that has exactly ("==") B or M in the diagnosis column will become part of the corresponding data set.The pull function is used to select a column in the cancer_sample data frame and transforms it into a vector, so it can be used in the test.

BS <-cancer_sample %>%
  arrange(smoothness_mean)%>%
  filter(diagnosis == "B")%>%
  pull(smoothness_mean)
head(BS)


MS <-cancer_sample %>%
  arrange(smoothness_mean)%>%
  filter(diagnosis == "M")%>%
  pull(smoothness_mean)
head(MS)

#We have to test if the data is normally distributed using the shapiro.test function, whose  null hypothesis is that the data is normally distributed. 

shapiro.test(BS)
shapiro.test(MS)


# Four different histograms were made in order to know which bin width is better for the data set. Bin intervals need to have enough distance to include the upper and lower spec limits and the min and max values, without having too many you can not see clearly the distribution.
#Bin width is used to change the with of the bins on the histogram, therefore the number of bins
#ggtitle is used to add a tittle to the graph, the name must be in quotation marks

H1 <- ggplot(cancer_sample, aes(x=smoothness_mean)) + 
  geom_histogram(aes(fill=diagnosis), alpha = 0.6, binwidth=0.05) +
  ggtitle("Bin size = 0.05")

H2 <- ggplot(cancer_sample, aes(x=smoothness_mean)) + 
  geom_histogram(aes(fill=diagnosis), alpha = 0.6, binwidth=0.01) +
  ggtitle("Bin size = 0.01")

H3 <- ggplot(cancer_sample, aes(x=smoothness_mean)) + 
  geom_histogram(aes(fill=diagnosis), alpha = 0.6, binwidth=0.005) +
  ggtitle("Bin size = 0.005")

H4 <- ggplot(cancer_sample, aes(x=smoothness_mean)) + 
  geom_histogram(aes(fill=diagnosis), alpha = 0.6, binwidth=0.001) +
  ggtitle("Bin size = 0.001")

#ggarrange from the ggpubr package allows the combination of different graphs in one figure for better visualization, you can set the number and rows and columns in which the graphs will be placed
Q2.2 <- ggarrange(H1, H2, H3, H4, 
          ncol = 2, nrow = 2)
print (Q2.2)

#The best histogram is the one with a bin width of 0.005, because it is small enough to see the distribution of the values without hiding important details.  Also is not too small so it doesn't generate a lot of noise, in which case a scatter plot would be best. 
Q2.3 <-H3
print(Q2.3)
```

**Analysis**

On the contrary of what the initial hypothesis was, the samples of benign tumors have more variability than the malign tumors. Benign tumors have a higher standard deviation and variance. Also the mean of the standard error of every sample has higher meaning that they have more variability in the smoothness on their nuclei between samples. 

The p-value of the Shapiro test is less than 0.05 implying that the distribution of the data is significantly different from normal distribution. Since the data is not normally distributed F test can't be used to compare variances between the smoothness in benign and malign tumors. The F-test in R can be performed using (var.test(x, y, alternative = ""))

Moreover in the histogram it can be seen in a more graphic way that there is more variability in the benign tumors than in the malign ones, there are more tall bars near the mean. Also, in general the bars are taller in the benign tumors but that is because the number of samples of benign tumors is bigger.

#### Third research questions
*Could it be possible to diagnose the tumor of a patience just by looking at the concavity of the nuclei?* 

The concavity of the nuclei refers to the severity of concave portions of the contour, so it is related with smoothness, the higher the concavity the less likely that cells look spherical like, healthy cells have spherical or oblong nucleus without severe concave portions. My hypothesis is that benign tumors will have lower concavity than malign tumors and that can be used as a factor to diagnose the patience using biopsies. To examine the data a summarize table was made to know the main static values of both groups and see if it is an evident difference between them. 

*Moreover three categorical values were created from the existing data of concavity*. Normal corresponds to nuclei whose concavity value is less than 0.1, which is closer to the shape of healthy cells. If the concavity is between 0.1 and 0.25 is labeled as high, since those values indicates that there is a clear problem with the cell. Higher than 0.25 were classified as very high, those cells have an aberrant nucleus and definitely are undergoing a tumorous process, on the contrary they should be undergoing apoptosis. 

The number of samples of each category separated by group will be shown in a table and in an *histogram with two geom layers and the exact number of samples in every group* to see if there is clear evidence that benign tumor have lower concavity values than malign.

```{r}
#To know the basic information about concavity a summarize table was created similar to the one done for the first research question but without grouping by diagnosis, just to know the range of values to generate the categories.
Q3.1 <- cancer_sample %>%
  arrange(concavity_mean)  %>%
  summarise(Mean_concavity = mean(concavity_mean),
            Min_concavity = min(concavity_mean),
            Median_concavity = median(concavity_mean),
            Max_concavity = max(concavity_mean)) %>%
              mutate(across(where(is.numeric), ~ round(., 3)))
Q3.1

# A tibble was created with just the variables of interest, diagnosis and concavity using th function select.
#Using the arrange function the tibble was reorder by the ascending values of concavity
#The numerical variable of concavity was divided in low, high, and very high categories using the mutate function to create a new column. 
#Case_when was used to categorize the values into 3 different categories, the true value is stated for the values that didn't meet the categories established before.
Q3.2 <- cancer_sample %>%
    select(diagnosis, concavity_mean) %>%
    arrange(concavity_mean) %>%
     mutate(concavity = case_when (concavity_mean <0.1 ~  "normal", 
                                   concavity_mean <0.25 ~  "high", 
                                   TRUE ~ "extremly high"))
Q3.2

#The count function is used view a summary of the frequencies of two categorical values, a tibble is created summarizing how many samples are in every category by group of diagnosis. 
Q3.3 <- count(Q3.2, diagnosis, concavity) 
head(Q3.3)


# A bar graph was created using the ggplot2 package to visualize the number of samples in every category, the function ..count.. is used instead of the y axis to the bar graph, been able to show the number of samples instead of the frequencies  
# Using geom_text the exact number of every category will appear in a label at the top of the bar graph 
Q3.4 <- ggplot(Q3.2, aes(concavity, ..count..)) +
   geom_bar(fill= "orange", alpha = 0.7, position = "dodge") +
   facet_wrap(~ diagnosis) +
   geom_text(aes(label = ..count..), stat = "count", vjust = 0.1)
print(Q3.4)

 
#A density plot was created to see if there is a clear difference between benign and malign tumors.
#Alpha transparency is used to visualize the intersections of the two groups 
Ex3.5 <- ggplot(cancer_sample, aes(concavity_mean)) + 
    geom_density(aes(fill = diagnosis), alpha=0.8)
print(Ex3.5)
```

**Analysis**

The concavity values goes from 0.00 to 0.427, therefore three categories were made to evaluate the severity of the concave points. Looking at the tibble Q3.3 it is evident that both groups have representation in the three categories so the hypothesis was not true, indeed benign tumors have a mayor percentage of "normal" concavity but there are some samples that are both in the high and very high categories, therefore the patients can't be diagnosed just by looking at the concavity of the nuclei. Looking at the bar graph in Q3.4 is clear that most samples are in the high and extremely high category is most probable that a malign tumors have nuclei with big concavity sections. 

Finally, analyzing the density plot there is visual evidence that both groups overlap and is impossible to separate them using the concavity values. The malign tumors have more variability than the benign ones, for that reason there are some samples that have values as low as zero and it wouldn't be correct to diagnose using just this variable. 

#### Forth research questions
*Is there a relationship between the symmetry of the nuclei and the compactness in malign tumors?* 

The compactness of the nuclei refers to  ratio between the perimeter and the area perimeter^2 / area - 1.0, so it is easier to obtain than symmetry. If there is a clear relationship between the symmetry and the compactness a model could be used to predict the value of the other one. Both values are significant because malign tumors tend to have less symmetry in their nuclei and bigger both area and perimeter.

A tibble was created to just see the symmetry and compactness of malign tumors. Moreover the size of the sample was computed by *counting the number of observations of malign tumors* to see if it is representative sample of the population. A graph was created to see the relationships of both variables using a scatter plot, *the scale of the y axis was changed to logarithmic* so the points in the scatter plot are not so overlap and it is easier to see a relationship. Latter on model fitting was used to actually see if there is a linear relationship between the two variables.

```{r}
#A tibble was create extracting just the information of malign tumors using the filter functions that returns juts the rows that contain M in the diagnosis column 
Q4.1 <- cancer_sample %>%
  rename(c("symmetry"="symmetry_mean", "compactness"="compactness_mean")) %>%
  select(diagnosis, symmetry, compactness) %>%
  arrange(symmetry)%>%
  filter(diagnosis == "M")
head(Q4.1)

#The count function gives back the number of observations in a categorical variable, since it is using the tibble created in Q4.1 it will yield the number of samples diagnosed as malign
Q4.2 <- count(Q4.1, diagnosis)
head(Q4.2)

# A scatter plot was created using geom_point.
# Geom_smooth aids the the visualization of patterns in the presence of over plotting, lm is used to see if there is a linear relationship
# The scale of the y axis was change to logarithmic using the scale_y_continuous function of scales package and is labeled using labels
#Code was created based in the one from: http://www.sthda.com/english/wiki/ggplot2-axis-scales-and-transformations#axis-transformations
Q4.3 <- ggplot(Q4.1, aes(symmetry, compactness)) + 
  geom_point() +
  geom_smooth(method = lm, formula = y ~ x) +
  scale_y_continuous(trans = log10_trans(),
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x))) +
    labs(title="Relationship between symmetry and compactness of the nuclei in malignt breast tumors", y ="Compactness", x = "Symmetry")
print(Q4.3)

#The lm function was used to obtain a linear model that shows the relationship between the two variables 
# Glance function was used from the broom package to see the a summary of the model fit
Q4.4 <- glance((lm(symmetry ~ compactness, Q4.1)))
print(Q4.4)
```
**Analysis**

The number malign samples is adequate for the study it has certain representativeness, more sample would be better considering the amount fo people with breast cancer but is good for a first approach. In the scatter plot it can be seen that in general there is a tendency in which with higher symmetry there is higher compactness, however there are some samples that do not follow that prediction. The smooth line presented in the graph shows that there is not a clear linear relationship.
By analyzing the results from the glance function is evident that the values don't have a linear relationship since the r.squared value is very low, meaning that the independent variable has not a linear tendency compared with the dependent variable. Therefore the model cannot be used to see predictions of one of the variables. 

## 1.3

As seen in the task 1.2 all questions were answered with the analysis done, however there is need for more information to validate those statements, for example a critical factor in cancer patients is the stage of the tumor, at higher stage the cells are more different than normal cells.  Knowing the stage of the tumors could be useful to separate the samples and have a more robust analyzes. To have  more reliable results more malign tumors would be needed, since the population of benign tumors is significantly bigger in this data set. Also an interest factor to analyze is the mutations related to benign and malign tumors on those samples, extracting DNA from the biopsy, however for that study more experiments would be needed. 
Moreover all the data was obtained thought machine learning, however it would be good that the information was validated by a pathologist to confirm that the findings are correct and there is no over or underestimation of the program, with this step the results obtained from the analysis can be validated 

Research question 1 yielded interesting results as it was seen that the is a clear difference between the average mean in the area of the nuclei, however there is wide variability in the samples so more information would be needed to validate the diagnosis.
Question 2 could be re-defined since not much information can be obtained from the variability of the sample, just to know that it cannot be quite useful for determining if the tumor is benign or malign. Another variable could be explore, like the perimeter. 
Research question 3 yielded different results from what I was expecting, since benign tumors also have very high concavity, the research question will be need to re-defined too include other variables as well so it would be possible to diagnose a tumor be just some values obtained from a biopsy.
The forth research questions could be re-defined, maybe different variables can be compared to create a model since the ones analyzed didn't have a linear relationship. Nevertheless, it is unclear if other models could fit to represent the relations between this two variables 

# Task 2: Tidy your data 

## 2.1
By selecting just 8 columns from cancer_samples and creating a tibble is evident that *the data is tidy* since, every header of a column is a variable that shows different aspects of the nuclei. There are not multiple variables in one column, of the samples, every row is an observation of a categorical variable, ether malign or benign and every cells has a numerical value. In this arrangement is easy to analyze the data, extracting certain variables and filtering specific observations, so it is possible to use the functions in the tidyverse package.
```{r}
#The first example of a tidy tibble was made by just selecting 8 columns from the original data set
T2.1<- cancer_sample %>%
  select(diagnosis, area_mean, perimeter_mean, smoothness_mean, concavity_mean, symmetry_mean, compactness_mean, texture_mean)
T2.1

#Another example of a tidy tibble is the one below, instead of having all the different observations individually they were summarized by group obtaining the mean along the two diagnosis for each numerical value using the summarize function  
T2.2<- cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(Mean_area= mean(area_mean), 
            Mean_perimeter= mean(perimeter_mean), 
            Mean_smoothness = mean(smoothness_mean),
            Mean_concavity= mean(concavity_mean),
            Mean_symmetry= mean(symmetry_mean), 
            Mean_compactness= mean(compactness_mean), 
            Mean_texture= mean(texture_mean))
T2.2
```
## 2.2
Since my data was tidy I untidy it too show how it would be messy to have it in and untidy form, if I try to untidy the T2.1 tidy tibble it would result in an error since values are not uniquely identified, for that reason the data will be untidy.

Furthermore, if I untidy the T2.2 tibble that contains the summarized values a tibble without error is created it is easier understand and to compared the two groups in the diagnostic variable, however it would be hard to work with it, needing to make some transformation to use ggplot2 and the tidyverse package, since there are variables stored in a variable (Measurement) 

Finally I tidy the untidy data to transform back to its original form in which analyzes can be done to answer the corresponding research questions.
```{r}
#The first tidy tibble was transformed using pivot_longer and pivot_wider from the tidyverse package 
#Pivot_longer increase the number of rows and decrease either number of column, so it will transpose the columns to rows, except diagnosis since it has a "-" in front of it, you need to establish the header of the new column with "name_to"
#Pivot_wider increase the number of columns and decrease the number of rows, so the names found in the column diagnosis will be placed as column headers and the values as cells.
 T2.3<- T2.1 %>%
               pivot_longer(cols = -diagnosis, names_to = 'Measrument')  %>% 
               pivot_wider(names_from = diagnosis, values_from = value)
T2.3

#Since the transformation above yielded an error I use the same code to transform the summarized tibble to show the result of untidy data
T2.4<- T2.2 %>%
               pivot_longer(cols = -diagnosis, names_to = "Measurement") %>% 
               pivot_wider(names_from = diagnosis, values_from = value)
T2.4

#To tidy the untidy data first all the columns headers (M and B) except measurements will become rows (observations) and all the names in the measurement column will become column headers, finally the values are obtained from the numerical variables in the tibble 
T2.5 <- T2.4 %>%
  pivot_longer(cols = c(-Measurement), names_to = "Diagnosis") %>% 
      pivot_wider(names_from = Measurement, values_from = value)
T2.5
```
## 2.3 

Based on my interest and the result that were obtained after the analysis I chose to continue with the next research questions 

*1. Could it be possible to diagnose the tumor of a patience just by looking at the concavity of the nuclei?*
*2. Is there a relationship between the area of the nuclei and the perimeter in malign tumors?*

Since the data set does not contain many categorical variables the categories I made from concavity could be useful to develop the exercises in Milestone 3, moreover I analyzed the other variables and all of them overlap between the two groups but the one with a clearer distinction is the concavity, also is a crucial element that pathologist see while analyzing biopsis.

Furthermore, I change one of my research questions so it would be possible to create a model that accurately creates predictions based on two variables. Both perimeter and area should be related be a linear model, however since the data was obtained by computational analysis maybe the model wont be perfect and I think is worth exploring it. Finally these two variables are the ones whose mean values grouped by diagnosis are more different, so analyzing them could give a good approach to determining the diagnose of a biopsy. 
```{r}
#The variables of interest were renamed so it will be easier to call them in a function
#The variables of interest were selected, dropping irrelevant columns
#It was arranged in ascending order according to the concavity
#It was grouped by diagnosis
#The mean of two groups were obtained
#Mutate was used to round to 3 decimals.
T2.6 <- cancer_sample %>%
  rename(c("concavity"="concavity_mean"))%>%
  select(diagnosis, concavity) %>%
  arrange(concavity)  %>%
  group_by(diagnosis)%>%
  summarise(Mean_concavity = mean(concavity)) %>%
  mutate(across(where(is.numeric), ~ round(., 3)))
T2.6

#The variables of interest were renamed so it will be easier to call them in a function
#The variables of interest were selected, dropping irrelevant columns
#It was arranged in ascending order according to the area
#It was filtered just the samples diagnosed as malign
#A scatterplot was created to see the relationship between the two variables 
T2.7<- cancer_sample %>%
  rename(c("area"="area_mean", "perimeter"="perimeter_mean")) %>%
  select(diagnosis, area, perimeter) %>%
  arrange(area)%>%
  filter(diagnosis == "M")%>%
  ggplot(aes(area, perimeter)) + 
  geom_point()
T2.7
  
```


